Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-02 forecastflows.tex
Some text for the end but not complete yet.

The misuse of models and misinterpretation of their outcomes (including by the authors) have multiple origins. One cause, well known, is the pressure to publish academic work, which can lower research standards. This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
This a growing concern, leading to increase in reproducibility and data sharing practices.
However, model development in ecology is still rarely transparent, which limits model understandability and prevents peers from properly identifying potential issues. A significant portion of scientific debate thus becomes lost in methodological considerations rather than advancing our ecological understanding.

>>>>

We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science. 
One well-known hurdle is that the pressure to publish academic work, which can lower research standards and make the added effort of this workflow seem ill-placed. This may be especially true for those who see science rewarded mostly through the shear number of publications. But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help. Further, growing concerns about how reproducible science is, especially in ecology where samples sizes are low and effects likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, as more value is placed on research that carefully developed, openly shared (for data and code) and acknowledges its uncertainty towards the aim of improving future data collection and model development (see Fig).

% This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues. Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations, but we believe our workflow provides an tractable step to fixing this. By focusing on model development more tightly tied to ecological expertise, we ague this workflow should broader the community that contributes to model development. As ecologists are increasingly expanding their computational toolkits, many field, and lab and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological, and---most importantly---to interrogate them.

Advancing ecology to where most researchers see models built more flexibly from their understanding of their ecological systems will not happen rapidly without a major shift in training, however. Much of ecology still divides the world into training for those focused on several groups. First, those who will conduct field and lab studies often learn a suite of particular tests matched to particular experiment designs or to simply match their predictor and response data to a flow diagram of variable types (add EXAMPLE and reference, something like: categorical $y$ and binary $y$ means ... I dunno, chi square?) and are left adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is. Instead, they are expected to collaborate with others when they need more complex models, which is a second group of ecological researchers---those who focus on complex statistical models and were thus trained more in model development, but often for highly specific applications (e.g., wildlife population estimates) where they may rarely build totally new generative models. Separate from these groups, process-based modelers do their own thing. Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data. While specialization is valuable, we argue the fundamental training in ecology has overly-siloed these groups and prevented more rapid progress.

Training all ecologists in our proposed workflow would break down barriers between different groups of ecologists today with likely major benefits for science. Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in our workflow would focus on learning to generate questions and then models, and then to simulate data from them. This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them. Through this and retrodictive checks most ecologists would more easily think through what parameters are most critical to their question and or aim (e.g., management) and also gain a much stronger connection to the level of uncertainty in many of ecological estimates. Empirical ecologists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models. Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting. 
% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches to better forecasts or something.

This type of workflow-focused training would make forecasting from models far more tractable ...
